---
title: "PostgreSQL and SQLite side by side"
author: "Richard Aubrey White"
date: "2026-08-05"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PostgreSQL and SQLite side by side}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`DBTable_v9` takes one `dbconfig` list and does the rest. Change the list and the same table
definition runs against a different engine. This vignette puts the PostgreSQL list and the SQLite
list next to each other, then lists what a user must know about the differences.

None of the code below runs. It reaches a database, and the build machine has none. Copy it into
your own script.

`vignette("csdb", package = "csdb")` is the worked example that does run, on SQLite.

## Two configurations

PostgreSQL reads nine settings. This example sets eight of them. The ninth is
`role_create_table`, which makes `create_table` issue `SET ROLE` before it creates a table.
This example reads the credentials from the environment itself, so that they stay out of
scripts and out of version control.

```{r postgres-config, eval = FALSE}
cfg_postgres <- list(
  driver = "PostgreSQL Unicode",
  server = "localhost",
  port = 5432,
  db = "mydb",
  schema = "public",
  user = Sys.getenv("DB_USER"),
  password = Sys.getenv("DB_PASSWORD"),
  sslmode = "require"
)
```

SQLite reads two. `db` is the path to the file, which `DBConnection_v9` creates on first connect,
along with any missing parent directory. The other seven settings are accepted and ignored, so a
configuration built from the same environment variables still works.

```{r sqlite-config, eval = FALSE}
cfg_sqlite <- list(
  driver = "SQLite",
  db = "~/data/mydb.sqlite"
)
```

csdb matches the driver string case-insensitively for SQLite: `"sqlite"`, `"SQLite"` and
`"SQLITE"` all select it. csdb matches `"PostgreSQL Unicode"` exactly, because it MUST equal an
entry in `odbcinst.ini`.

## The same table against each

The table definition does not change. Only the `dbconfig` argument does.

```{r same-table, eval = FALSE}
field_types <- c(
  "location_code" = "TEXT",
  "date" = "DATE",
  "cases_n" = "INTEGER",
  "cases_pr100000" = "DOUBLE"
)

tab_postgres <- csdb::DBTable_v9$new(
  dbconfig = cfg_postgres,
  table_name = "cases",
  field_types = field_types,
  keys = c("location_code", "date"),
  indexes = list("ind1" = c("location_code", "date"))
)

tab_sqlite <- csdb::DBTable_v9$new(
  dbconfig = cfg_sqlite,
  table_name = "cases",
  field_types = field_types,
  keys = c("location_code", "date"),
  indexes = list("ind1" = c("location_code", "date"))
)

d <- data.table::data.table(
  location_code = "county_nor03",
  date = as.Date("2024-01-01"),
  cases_n = 12L,
  cases_pr100000 = 2.4
)

tab_postgres$insert_data(d)
tab_sqlite$insert_data(d)
```

## What differs

| | PostgreSQL | SQLite |
|---|---|---|
| `schema` | Read. Identifiers are `schema.table`. | Ignored entirely. SQLite has no schemas, so identifiers are the bare table name. |
| Primary key | Added after creation with `ALTER TABLE ... ADD CONSTRAINT`. | Inlined in `CREATE TABLE`, and every key column is `NOT NULL`. It cannot be added later: SQLite has no `ALTER TABLE ... ADD CONSTRAINT`, so `$add_constraint()` is a no-op. |
| An unrecognised `field_types` value | Passed through to the engine, which decides. | Rejected before any SQL is emitted, naming the column and the type. |
| `get_table_names_and_info()` row count | The `reltuples` estimate the planner holds. | An exact `COUNT(*)`. |
| `get_table_names_and_info()` sizes | Real total, data and index sizes in GB. | `NA` in all three size columns. |
| External client binary | `psql` must be on `PATH`. Bulk loads are staged to a CSV file and shelled out. | None: no `psql` and no `bcp`. Rows go straight through `DBI::dbAppendTable()`. |
| `$keep_rows_where()` | Copies the surviving rows to a new table, drops the old one, renames. | A single `DELETE`, which leaves the primary key and the indexes in place. |

### Why the field types are closed

SQLite accepts any declared type name. `VARCHAR(100)`, `TEXT(100)` or a misspelling would all
create a table, with an unintended affinity and no warning. `csdb` therefore accepts six types
and rejects the rest:

```{r rejected-type, eval = FALSE}
csdb::DBTable_v9$new(
  dbconfig = cfg_sqlite,
  table_name = "cases",
  field_types = c("location_code" = "VARCHAR(100)", "cases_n" = "INTEGER"),
  keys = "location_code"
)$create_table()
# SQLite does not support the field type(s): location_code (VARCHAR(100)).
# The supported types are: TEXT, INTEGER, DOUBLE, BOOLEAN, DATE, DATETIME.
```

`DATE` and `DATETIME` are declared types rather than storage classes. `DBConnection_v9` opens the
file with `extended_types = TRUE`, which is what reads them back as `Date` and `POSIXct`. Without
that argument `2020-01-01` returns the integer `18262`.

### Why the primary key cannot be added later

`ALTER TABLE cases ADD CONSTRAINT ... PRIMARY KEY (...)` is `near "CONSTRAINT": syntax error` in
SQLite. The key MUST therefore exist from the first `CREATE TABLE`. Two consequences follow.
`$upsert_data()` needs that key, because `ON CONFLICT (<keys>)` needs a matching `PRIMARY KEY`
or `UNIQUE` constraint. And every delete path is a `DELETE` rather than a drop-and-rename, because
a dropped key could not be put back.

## Which one to choose

Use SQLite for tests, for examples, and for a single-writer pipeline on one machine. It is a file:
there is nothing to install, nothing to start, and nothing to authenticate against.

Use PostgreSQL when more than one process writes, or when the data outgrows one machine. Also use
it when the row counts get large enough that an exact `COUNT(*)` is too slow. A planner estimate
is then the honest answer.
