---
title: "Introduction to csdb"
author: "Richard Aubrey White"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to csdb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# error = FALSE is load-bearing, not decoration. knitr defaults it to TRUE,
# which renders a failure into the page and lets the build report success. An
# unexpected error below MUST stop `R CMD check`.
#
# The 6 chunks that demonstrate a raise set error = TRUE for themselves,
# because the message they print is the demonstration. Such a chunk passes
# whether it raises or not, so those raises are demonstrations and NOT
# assertions. See "What this vignette does and does not prove".
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)
```

## What csdb is for

`csdb` puts a large surveillance dataset into a database and takes it out again, without you
writing SQL. You describe a table once: its columns and their types, its primary key, and its
indexes. You then insert, upsert, delete and read rows through R methods. Before it writes, it
can check that your data matches an agreed column format, so a table cannot quietly change shape
between runs.

No chunk in this vignette needs a database server. Every chunk that reaches a database uses
SQLite, in a file that `tempfile()` names. `vignette("backends", package = "csdb")` puts a
PostgreSQL configuration and a SQLite configuration side by side.

### What this vignette does and does not prove

`R CMD check` re-runs every chunk. A chunk with `error = FALSE` turns the check red if it raises
unexpectedly. A chunk with `error = TRUE` passes whether it raises or not.

This vignette therefore guards against new unexpected errors. The expected errors and every
printed value are demonstrations, and the test suites assert them. See `tests/testthat/` in the
package source.

### Two classes, and the split between them

`DBConnection_v9` is the connection: the settings, plus `$connect()`, `$disconnect()` and
`$autoconnection`. `DBTable_v9` is one table: its field types, its keys, its indexes, and every
read and write method.

You do not have to build the connection first. `DBTable_v9$new()` takes the same settings as a
`dbconfig` list, builds its own `DBConnection_v9`, and keeps it at `$dbconnection`. Build a
`DBConnection_v9` yourself when you want a connection without a table.

Neither constructor touches the database. Both return an object that is not connected:

```{r}
con <- csdb::DBConnection_v9$new(driver = "SQLite", db = tempfile(fileext = ".sqlite"))
con$is_connected()

tab <- csdb::DBTable_v9$new(
  dbconfig = list(
    driver = "PostgreSQL Unicode",
    server = "localhost",
    port = 5432,
    db = "mydb",
    schema = "public",
    user = "u",
    password = "p"
  ),
  table_name = "cases",
  field_types = c("location_code" = "TEXT", "date" = "DATE", "cases_n" = "INTEGER"),
  keys = c("location_code", "date")
)
tab$dbconnection$is_connected()
```

The second block names a PostgreSQL server that is not running, and still returns an object. The
first method that reaches the database is what connects.

### The validators, and one gap

A validator is a function you hand to `DBTable_v9$new()`. `csdb` exports six:

```{r}
sort(grep("^validator_", getNamespaceExports("csdb"), value = TRUE))
```

The `_blank` pair accepts anything. The other four check two csverse surveillance formats,
`csfmt_rts_data_v1` and `csfmt_rts_data_v2`. A `_field_types` validator runs once, inside
`$new()`. A `_field_contents` validator runs inside `$insert_data()` and `$upsert_data()`.

**There is no `csfmt_rts_data_v3` validator.** That is a current limitation, and it matters now.
`cstidy` marks v1 and v2 deprecated, and points new work at `set_csfmt_rts_data_v3()`.
`csalert`'s pipeline ends in `ens_collapse(heal = TRUE)`, which returns a `csfmt_rts_data_v3`. So
today's analysis output has no validator here that matches it. You can still store it: pass the
`_blank` pair, or a function of your own. `csdb` then checks nothing about the columns.

### The three backends

| Backend | Connects through | Needs a server |
|---|---|---|
| PostgreSQL | `odbc`, driver `"PostgreSQL Unicode"` | Yes |
| Microsoft SQL Server | `odbc`, driver `"ODBC Driver 17 for SQL Server"` | Yes |
| SQLite | `RSQLite`, where `db` is a file path | No |

One `dbconfig` list picks the backend, and nothing else in your code changes. `csdb` reads no
environment variable itself. To keep a password out of a script, read the variable yourself and
pass the result to `DBTable_v9$new()`.

### Where csdb sits

`csdb` is the storage layer under the csverse surveillance stack. It imports `csutil`. Its
validators are written against `cstidy`'s data formats, by column name and type. No `csdb` code
calls `cstidy` or `csalert`, and neither is a dependency.

## A worked example

The block below creates a table object, clears it, and inserts the bundled
`nor_covid19_cases_by_time_location` dataset. The file is new, so `$drop_all_rows()` creates the
table and its index first, and then reports that it removed 0 rows. `$tbl()` returns a lazy
dbplyr reference.

```{r}
library(data.table)

db_file <- tempfile(fileext = ".sqlite")

dbconnection <- csdb::DBConnection_v9$new(
  driver = "SQLite",
  db = db_file
)
dbconnection
dbconnection$connect()
dbconnection$connection
dbconnection$disconnect()
class(dbconnection$connection)

dbtable <- csdb::DBTable_v9$new(
  dbconfig = list(
    driver = "SQLite",
    db = db_file
  ),
  table_name = "anon_test",
  field_types = c(
    "granularity_time" = "TEXT",
    "granularity_geo" = "TEXT",
    "country_iso3" = "TEXT",
    "location_code" = "TEXT",
    "border" = "INTEGER",
    "age" = "TEXT",
    "sex" = "TEXT",
    "isoyear" = "INTEGER",
    "isoweek" = "INTEGER",
    "isoyearweek" = "TEXT",
    "season" = "TEXT",
    "seasonweek" = "DOUBLE",
    "calyear" = "INTEGER",
    "calmonth" = "INTEGER",
    "calyearmonth" = "TEXT",
    "date" = "DATE",
    "covid19_cases_testdate_n" = "INTEGER",
    "covid19_cases_testdate_pr100000" = "DOUBLE"
  ),
  keys = c(
    "granularity_time",
    "location_code",
    "date",
    "age",
    "sex"
  ),
  indexes = list(
    "ind1" = c("granularity_time", "location_code", "date")
  ),
  validator_field_types = csdb::validator_field_types_blank,
  validator_field_contents = csdb::validator_field_contents_blank
)
dbtable$drop_all_rows()
dbtable$insert_data(csdb::nor_covid19_cases_by_time_location)
dbtable$tbl()
dbtable$disconnect()
```

## Who owns the connection

`DBTable_v9$new()` takes an optional `dbconnection` argument. That argument decides who closes
the connection.

Pass nothing and the table builds its own `DBConnection_v9`. The table owns it, and
`$disconnect()` closes it. Pass one in and the table borrows it. The caller keeps ownership, and
the borrower's `$disconnect()` closes nothing.

Ownership exists because one connection can serve many tables. A caller that builds many tables
against one database then holds one connection, and not one connection per table.

```{r}
# A table that builds its own connection owns it.
owner <- csdb::DBTable_v9$new(
  dbconfig = list(driver = "SQLite", db = tempfile(fileext = ".sqlite")),
  table_name = "anon_own",
  field_types = c("x" = "TEXT"),
  keys = "x"
)
owner$connect()
owner$dbconnection$is_connected()

# The owner closes its own connection.
owner$disconnect()
owner$dbconnection$is_connected()
```

```{r}
shared_file <- tempfile(fileext = ".sqlite")
shared <- csdb::DBConnection_v9$new(driver = "SQLite", db = shared_file)

borrower <- csdb::DBTable_v9$new(
  dbconfig = list(driver = "SQLite", db = shared_file),
  table_name = "anon_borrow",
  field_types = c("x" = "TEXT"),
  keys = "x",
  dbconnection = shared
)
borrower$connect()
shared$is_connected()

# The borrower does not own the connection, so this closes nothing.
borrower$disconnect()
shared$is_connected()

# The owner closes it.
shared$disconnect()
shared$is_connected()
```

## A connection does not survive a fork

A connection belongs to the process that opened it. After a fork, the parent and the child both
hold one socket. An inherited PostgreSQL connection can then return wrong rows, and report no
error. Measured against the NorSySS server on 2026-08-14. A child asked for `select 4` and read
3. The parent asked for `select 999` and read 2.

`DBI::dbIsValid()` cannot see this. It reports TRUE on an inherited handle, so a reconnect that
tests validity never fires.

`DBConnection_v9` records the process that opened each connection instead. It drops any
connection whose recorded process is not the current one. The block below changes that recorded
process rather than forking, because `parallel::mcparallel()` does not exist on Windows.

```{r}
db <- csdb::DBConnection_v9$new(driver = "SQLite", db = tempfile(fileext = ".sqlite"))
db$connect()
handle <- db$connection

# What DBI sees, before and after the connection becomes another process's.
DBI::dbIsValid(handle)
db$.__enclos_env__$private$pconnection_pid <- Sys.getpid() + 1L
DBI::dbIsValid(handle)

# What csdb sees.
db$is_connected()
is.null(db$connection)

# $autoconnection opens a handle for this process.
identical(db$autoconnection, handle)

# The inherited handle stays open. Closing it would close the other
# process's socket.
DBI::dbIsValid(handle)
db$disconnect()
```

The example reaches into `$.__enclos_env__$private`, which your own code MUST NOT do.

A fork writes nothing into that field. The child inherits the recorded opener process ID
unchanged, and it is the child's own process ID that differs. The example holds the current
process ID fixed and moves the recorded one, which produces the same mismatch from the other
side.

## Two tables that declare one index name

An index name is unique per schema on PostgreSQL, and unique per file on SQLite. Two tables that
both declare `ind1` therefore ask for one name. `CREATE INDEX IF NOT EXISTS` answers a name that
is already taken with a notice, and not with an error. The first table won the name, and every
later table silently got no index.

The names in `indexes = list("ind1" = ...)` are logical names. Each one reaches the database
under a physical name that carries the table identity. Two tables can now declare `ind1` and both
get an index.

```{r}
index_file <- tempfile(fileext = ".sqlite")

new_table <- function(nm) {
  csdb::DBTable_v9$new(
    dbconfig = list(driver = "SQLite", db = index_file),
    table_name = nm,
    field_types = c("x" = "TEXT", "n" = "INTEGER"),
    keys = "x",
    indexes = list("ind1" = "n")
  )
}

alpha <- new_table("anon_alpha")
beta <- new_table("anon_beta")
alpha$add_indexes()
beta$add_indexes()

# Two tables, one declared name, two indexes.
DBI::dbGetQuery(
  alpha$dbconnection$autoconnection,
  "SELECT tbl_name, name FROM sqlite_master
    WHERE type = 'index' AND name NOT LIKE 'sqlite\\_%' ESCAPE '\\'
    ORDER BY tbl_name"
)
```

`add_indexes()` reads the catalogue after each create. A statement that returns without an error
proves nothing about which table holds the name. The method raises when the index is absent from
this table, and when it covers columns other than the declared ones. That check is defined for
SQLite and for PostgreSQL, and for no other backend.

```{r}
alpha$disconnect()
beta$disconnect()
```

## Emptying a table before you write to it

`$drop_all_rows_and_then_insert_data()` and `$drop_all_rows_and_then_upsert_data()` empty the
table and then write. Both check `newdata` before they drop a row, and both answer these four
cases the same way.

```{r}
# Every table below starts with the same three sentinel rows. A row that
# survives a rejected call proves the guard ran before the drop.
d <- data.table::data.table(x = c("a", "b", "c"), n = 1:3)

seeded_table <- function(nm, validator) {
  tab <- csdb::DBTable_v9$new(
    dbconfig = list(driver = "SQLite", db = tempfile(fileext = ".sqlite")),
    table_name = nm,
    field_types = c("x" = "TEXT", "n" = "INTEGER"),
    keys = "x",
    validator_field_contents = validator
  )
  tab$insert_data(d)
  tab
}

# Refuses any frame holding a value of n that is not positive. A zero-row
# frame passes it, because all(integer(0) > 0) is TRUE.
positive_only <- function(data) all(data$n > 0)

# Refuses a zero-row frame, and accepts everything else.
nonempty_only <- function(data) nrow(data) > 0

bad <- data.table::data.table(x = c("d", "e"), n = c(5L, -1L))
```

**1. A `NULL` raises, and the table keeps every row.** `insert_data()` returns early on a `NULL`,
so its validator never sees one. Without this check the table emptied and nothing raised.

```{r, error = TRUE}
tab_null <- seeded_table("anon_guard_null", positive_only)
tab_null$nrow()

tab_null$drop_all_rows_and_then_insert_data(NULL)
tab_null$nrow()
```

**2. A zero-row frame that the validator refuses raises, and the table keeps every row.**

```{r, error = TRUE}
tab_zero_bad <- seeded_table("anon_guard_zero_bad", nonempty_only)
tab_zero_bad$nrow()

tab_zero_bad$drop_all_rows_and_then_insert_data(d[0])
tab_zero_bad$nrow()
```

**3. A zero-row frame that the validator accepts empties the table, and raises nothing.** That is
how `cs9::DBPartitionedTableExtended_v9` clears a partition.

```{r}
tab_zero_ok <- seeded_table("anon_guard_zero_ok", positive_only)
tab_zero_ok$nrow()

tab_zero_ok$drop_all_rows_and_then_insert_data(d[0])
tab_zero_ok$nrow()
```

**4. A frame with rows reaches the validator before the drop.** A refused frame raises, and every
sentinel row is still there afterwards.

```{r, error = TRUE}
tab_rows <- seeded_table("anon_guard_rows", positive_only)
tab_rows$nrow()

tab_rows$drop_all_rows_and_then_insert_data(bad)
tab_rows$tbl() |> dplyr::collect()
```

An accepted frame empties the table and takes the write.

```{r}
tab_rows$drop_all_rows_and_then_insert_data(d[1:2])
tab_rows$nrow()
```

### The upsert method answers the same four cases

`$drop_all_rows_and_then_upsert_data()` runs the same guard, in the same four cases and the same
order.

```{r, error = TRUE}
# 1. A NULL raises.
tab_up <- seeded_table("anon_guard_upsert", positive_only)
tab_up$drop_all_rows_and_then_upsert_data(NULL)
tab_up$nrow()
```

```{r, error = TRUE}
# 2. A zero-row frame the validator refuses raises, and every sentinel row
#    stays. This is the case that tells "returns early on zero rows" apart
#    from "validates zero rows too".
tab_up_zero <- seeded_table("anon_guard_upsert_zero", nonempty_only)
tab_up_zero$drop_all_rows_and_then_upsert_data(d[0])
tab_up_zero$tbl() |> dplyr::collect()
```

```{r}
# 3. A zero-row frame the validator accepts empties the table.
tab_up$drop_all_rows_and_then_upsert_data(d[0])
tab_up$nrow()
```

```{r, error = TRUE}
# 4. A frame with rows reaches the validator before the drop. A refused frame
#    leaves every sentinel row in place.
tab_up_rows <- seeded_table("anon_guard_upsert_rows", positive_only)
tab_up_rows$drop_all_rows_and_then_upsert_data(bad)
tab_up_rows$tbl() |> dplyr::collect()
```

```{r}
# An accepted frame empties the table and takes the write.
tab_up_rows$drop_all_rows_and_then_upsert_data(d[1:2])
tab_up_rows$nrow()
```

The guard covers what `csdb` can see before the drop. It is not a transaction. A write that fails
after the truncation still leaves the table empty.

```{r, include = FALSE}
tab_null$disconnect()
tab_zero_bad$disconnect()
tab_zero_ok$disconnect()
tab_rows$disconnect()
tab_up$disconnect()
tab_up_zero$disconnect()
tab_up_rows$disconnect()
```
